---
title: "A Practical Introduction to Quarto"
subtitle: "Create reproducible documents with text, code, tables, and figures"
author:
  - name: Lucas Toshio Ito
    orcid: 0000-0001-9713-8687
    email: lucas.toshio@unifesp.br
    affiliations:
      - name: Federal University of São Paulo (UNIFESP)
date: last-modified
toc: true
toc-location: left
toc-title: Contents
number-sections: true
highlight-style: github
format:
  html:
    theme:
      light: flatly
      dark: darkly
    code-fold: false
    code-tools: true
    embed-resources: true
    df-print: paged
css: styles.css
execute:
  echo: true
  warning: false
  message: false
---

[View the source of this tutorial](https://github.com/lcstoshio/Lcstoshio.github.io/blob/main/pages/Tutorial_Quarto/index.qmd)

# What is Quarto?

[Quarto](https://quarto.org/) is an open-source publishing system for creating reproducible documents, presentations, websites, and books. A Quarto file combines:

- Markdown text;
- executable code in R, Python, Julia, or Observable;
- results such as figures and tables;
- document settings in a YAML header.

Quarto files use the `.qmd` extension. This tutorial creates an HTML document with R, but the same ideas apply to other languages and output formats.

::: {.callout-tip}
## The basic workflow

Write in a `.qmd` file, render it, inspect the output, and repeat. The source remains readable and records how every result was produced.
:::

# Install the tools

Install:

1. [Quarto](https://quarto.org/docs/get-started/);
2. [R](https://cran.r-project.org/);
3. [RStudio](https://posit.co/download/rstudio-desktop/) or another editor with Quarto support.

Confirm that Quarto is available from a terminal:

```{bash}
#| eval: false

quarto --version
```

In RStudio, choose **File > New File > Quarto Document**. From a terminal, you can also create an empty file named `index.qmd`.

# Your first document

A minimal Quarto document contains a YAML header followed by Markdown content:

````markdown
---
title: "My first Quarto document"
author: "Your name"
format: html
---

# Introduction

This is a reproducible document.

```{r}
summary(mtcars$mpg)
```
````

Save the file and render it with the **Render** button in RStudio or with:

```{bash}
#| eval: false

quarto render index.qmd
```

The command creates `index.html` in the same directory unless another output location is configured.

# The YAML header

YAML controls document-level settings. It begins and ends with `---`.

```yaml
---
title: "Analysis report"
subtitle: "A reproducible example"
author: "Your name"
date: today
toc: true
number-sections: true
format:
  html:
    theme: cosmo
    code-fold: true
execute:
  warning: false
  message: false
---
```

Indentation matters in YAML. Use spaces, keep nested settings aligned, and do not use tabs.

Useful options include:

| Option | Purpose |
|---|---|
| `toc: true` | Add a table of contents |
| `number-sections: true` | Number headings |
| `code-fold: true` | Let readers show or hide code |
| `embed-resources: true` | Produce a self-contained HTML file |
| `execute: echo: false` | Run code without displaying it |

# Write with Markdown

Markdown provides lightweight formatting.

```markdown
# Level-one heading
## Level-two heading

**Bold text** and *italic text*

- unordered item
- another item

1. first step
2. second step

[Quarto documentation](https://quarto.org/docs/guide/)

`inline code`
```

## Callouts

Callouts highlight information without custom HTML.

```markdown
::: {.callout-warning}
## Check your data

Never overwrite the raw data during cleaning.
:::
```

::: {.callout-warning}
## Check your data

Never overwrite the raw data during cleaning.
:::

## Tabsets

Add `.panel-tabset` to a heading when readers should switch between related views.

````markdown
## Results {.panel-tabset}

### Summary

Summary text goes here.

### Code

```{r}
summary(mtcars)
```
````

# Add executable R code

An R code cell starts with ```` ```{r} ```` and ends with ```` ``` ````. In RStudio, insert one with **Ctrl+Alt+I** on Windows or **Cmd+Option+I** on macOS.

```{r}
# A cell can create objects and print results.
mean_mpg <- mean(mtcars$mpg)
mean_mpg
```

Objects created in one cell are available to later cells during the same render.

## Cell options

Cell options begin with `#|` and control that specific cell.

```{r}
#| label: mpg-summary
#| echo: true
#| output: true

summary(mtcars$mpg)
```

Frequently used options:

| Cell option | Effect |
|---|---|
| `#| echo: false` | Hide the source code |
| `#| output: false` | Hide printed output |
| `#| include: false` | Hide both code and output |
| `#| eval: false` | Display code without running it |
| `#| warning: false` | Hide warnings |
| `#| fig-width: 7` | Set figure width in inches |

## Inline R code

Use inline code for values that should update automatically. For example, this document calculates the mean fuel economy as `r round(mean_mpg, 1)` miles per gallon.

The source is:

```markdown
The mean is `r round(mean_mpg, 1)` miles per gallon.
```

# Tables

Any printed data frame appears as a table. For a compact publication-style table, use `knitr::kable()`.

```{r}
#| label: tbl-cars
#| tbl-cap: "Average fuel economy and horsepower by cylinder count"

car_summary <- mtcars |>
  transform(cyl = factor(cyl)) |>
  aggregate(cbind(mpg, hp) ~ cyl, data = _, FUN = mean)

knitr::kable(
  car_summary,
  digits = 1,
  col.names = c("Cylinders", "Mean MPG", "Mean horsepower")
)
```

Refer to the result as @tbl-cars. A cell label beginning with `tbl-` and a `tbl-cap` option create a numbered, cross-referenceable table.

# Figures

R graphics are inserted at the location of the code cell.

```{r}
#| label: fig-mpg-weight
#| fig-cap: "Heavier cars generally have lower fuel economy."
#| fig-alt: "Scatterplot of car weight against fuel economy with a downward fitted line."
#| fig-width: 7
#| fig-height: 5

plot(
  mtcars$wt,
  mtcars$mpg,
  pch = 19,
  col = "#2878B5",
  xlab = "Weight (1,000 lb)",
  ylab = "Fuel economy (mpg)"
)
abline(lm(mpg ~ wt, data = mtcars), col = "#C82423", lwd = 2)
```

As shown in @fig-mpg-weight, labels beginning with `fig-` can be referenced anywhere in the document.

# Citations and references

Add a bibliography file to the YAML header:

```yaml
bibliography: references.bib
```

Then cite an entry by its key:

```markdown
Quarto supports citations in many output formats [@citation-key].
```

Quarto automatically creates the reference list. Citation managers such as Zotero can export `.bib` files.

# Projects and file organization

For a single report, a small structure is enough:

```text
my-report/
├── index.qmd
├── data/
│   └── observations.csv
├── figures/
└── references.bib
```

For a website or book, create a Quarto project with `_quarto.yml`:

```yaml
project:
  type: website

website:
  title: "My project"
  navbar:
    left:
      - href: index.qmd
        text: Home
      - href: analysis.qmd
        text: Analysis

format:
  html:
    theme: cosmo
```

Render every file in the project with:

```{bash}
#| eval: false

quarto render
```

# Reproducibility practices

::: {.callout-important}
## A reproducible document should render from a clean session

Do not rely on objects that exist only in your interactive R environment.
:::

Recommended habits:

1. Keep raw data unchanged in a dedicated folder.
2. Use relative paths rather than computer-specific absolute paths.
3. Load every required package in the document.
4. Set a random seed when an analysis uses randomness.
5. Record package versions with `renv` when reproducibility matters.
6. Render from a fresh R session before sharing.
7. Track source files with Git; avoid committing temporary files.

```{r}
#| eval: false

set.seed(2026)
sample(1:100, size = 5)
```

# Preview, render, and share

Use live preview while writing:

```{bash}
#| eval: false

quarto preview index.qmd
```

Create the final output with:

```{bash}
#| eval: false

quarto render index.qmd
```

You can share the generated HTML file directly when `embed-resources: true` is enabled. Quarto can also publish projects to GitHub Pages, Quarto Pub, Netlify, or another web host.

# A reusable report template

````markdown
---
title: "Project report"
author: "Your name"
date: today
toc: true
format:
  html:
    theme: cosmo
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

set.seed(2026)
```

# Question

State the research question.

# Data

Describe the source, variables, and exclusions.

# Analysis

Show the reproducible analysis.

# Results

Present the main tables and figures.

# Conclusions

Summarize the answer, limitations, and next steps.
````

# Exercises

1. Create a new Quarto document and add your name, a table of contents, and numbered sections.
2. Add a code cell that calculates the mean and standard deviation of `iris$Sepal.Length`.
3. Create a table summarizing mean sepal length by species.
4. Create and cross-reference a figure of sepal length versus petal length.
5. Add a callout explaining one limitation of the analysis.
6. Render the document from a fresh R session and share the resulting HTML file.

# Next steps

- [Quarto guide](https://quarto.org/docs/guide/)
- [Quarto gallery](https://quarto.org/docs/gallery/)
- [Markdown basics](https://quarto.org/docs/authoring/markdown-basics.html)
- [Computations with R](https://quarto.org/docs/computations/r.html)
- [Publishing with Quarto](https://quarto.org/docs/publishing/)
