---
title: "A Practical Introduction to Quarto"
subtitle: "A short guide to reproducible documents with R"
author:
  - name: Lucas Toshio Ito
    orcid: 0000-0001-9713-8687
    email: lucas.toshio@unifesp.br
    affiliations:
      - name: Universidade Federal de São Paulo (UNIFESP)
date: last-modified
toc: true
toc-location: left
toc-title: Contents
number-sections: false
highlight-style: github
format:
  html:
    theme:
      light: flatly
      dark: darkly
    code-tools: true
    embed-resources: true
execute:
  warning: false
  message: false
---

[View the source of this tutorial](https://github.com/lcstoshio/Lcstoshio.github.io/blob/main/pages/Tutorial_Quarto/index.qmd)

# What is Quarto?

[Quarto](https://quarto.org/) combines Markdown text, executable code, figures, and tables in one reproducible document. Quarto files use the `.qmd` extension and can produce HTML, PDF, Word, presentations, websites, and books.

![](https://w0.peakpx.com/wallpaper/236/769/HD-wallpaper-moon-cat-pretty-wonderful-stunning-adorable-clouds-sweet-nice-art-kitty-black-sky-abstract-cat-cute-fire-awesome-moonlight-cats-beautiful-animal-draw-moon-light-blue-animals.jpg){fig-alt="An illustrated black cat sitting under a bright moon in a blue night sky." width="65%" fig-align="center"}

# Create a document

Install [Quarto](https://quarto.org/docs/get-started/), R, and RStudio. In RStudio, choose **File > New File > Quarto Document**.

A minimal document looks like this:

````markdown
---
title: "My first Quarto document"
author: "Your name"
format: html
---

# Introduction

This text is written with Markdown.

```{r}
summary(mtcars$mpg)
```
````

The section between `---` markers is the YAML header. It controls the title, author, output format, table of contents, theme, and other document-level options. YAML indentation must use spaces, not tabs.

# Write with Markdown

```markdown
# Main heading
## Subheading

**Bold text** and *italic text*

- first item
- second item

1. first step
2. second step

[Quarto documentation](https://quarto.org/docs/guide/)

`inline code`
```

Use callouts to highlight important information:

```markdown
::: {.callout-tip}
Keep raw data unchanged and perform cleaning in code.
:::
```

::: {.callout-tip}
Keep raw data unchanged and perform cleaning in code.
:::

# Add R code

Insert an R cell with **Ctrl+Alt+I** on Windows or **Cmd+Option+I** on macOS.

```{r}
mean_mpg <- mean(mtcars$mpg)
mean_mpg
```

Objects created in one cell are available to later cells during the same render. Inline code inserts a calculated value directly into a sentence: the mean fuel economy in `mtcars` is `r round(mean_mpg, 1)` miles per gallon.

Cell options begin with `#|`:

```{r}
#| label: mpg-summary
#| echo: true

summary(mtcars$mpg)
```

| Option | Effect |
|---|---|
| `#| echo: false` | Run the code but hide it |
| `#| output: false` | Hide printed output |
| `#| include: false` | Hide both code and output |
| `#| eval: false` | Display code without running it |
| `#| fig-width: 7` | Set figure width in inches |

# Tables and figures

Use `knitr::kable()` for a simple table:

```{r}
#| label: tbl-cars
#| tbl-cap: "Average fuel economy by cylinder count"

car_summary <- aggregate(mpg ~ cyl, data = mtcars, FUN = mean)
knitr::kable(
  car_summary,
  digits = 1,
  col.names = c("Cylinders", "Mean MPG")
)
```

Use a `tbl-` label to reference @tbl-cars. Figures work the same way with a `fig-` label and caption:

```{r}
#| label: fig-mpg-weight
#| fig-cap: "Heavier cars generally have lower fuel economy."
#| fig-alt: "Scatterplot of car weight and fuel economy with a downward fitted line."

plot(
  mtcars$wt,
  mtcars$mpg,
  pch = 19,
  col = "#2878B5",
  xlab = "Weight (1,000 lb)",
  ylab = "Fuel economy (mpg)"
)
abline(lm(mpg ~ wt, data = mtcars), col = "#C82423", lwd = 2)
```

See @fig-mpg-weight.

# Render and share

Click **Render** in RStudio or run:

```{bash}
#| eval: false

quarto render index.qmd
```

For a self-contained HTML file, add this to the YAML header:

```yaml
format:
  html:
    embed-resources: true
```

Before sharing, render from a fresh R session. Load every required package in the document, use relative file paths, set a random seed when needed, and keep raw data unchanged.

# Useful links

- [Quarto guide](https://quarto.org/docs/guide/)
- [Markdown basics](https://quarto.org/docs/authoring/markdown-basics.html)
- [Computations with R](https://quarto.org/docs/computations/r.html)
- [Publishing with Quarto](https://quarto.org/docs/publishing/)
